Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-08 paper/manuscript.tex

As discussed above, the situation for binding is more complex and requires special care. The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor. Doing otherwise would fail to specify the bound macrostate of the complex for which we seek the free energy. Moreover, the indicator function prescribes that the ligand remains in the binding site region throughout the alchemical process, including when it is uncoupled from the receptor. Consequently, alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and return undefined free energy estimates. Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

>>>>

As discussed above, the situation for binding is more complex and requires special care. The first important aspect is that binding free energy calculations must be performed with binding site restraints that implement the binding site indicator function $I(\zeta)$ by forcing the ligand to occupy the binding site even when it is decoupled from the receptor. Doing otherwise would fail to specify the bound macrostate of the complex and return undefined binding free energy estimates. However, in practice, calculations lacking binding site restraints often return reliable binding free energy estimates corresponding to the limited sampled binding site region explored during molecular dynamics. Nevertheless, longer calculations could display persistent drift of the estimate as new regions of conformational space are sampled. A more serious consequence of neglecting binding site restraints is the possibility of the ligand leaving the binding site when it is not coupled to the receptor. Such occurrence indicates a serious violation of the theory. Alchemical binding free energy simulations where the ligand leaves the binding site are not set up correctly and can return grossly inaccurate binding free energy estimates.

Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.
